%
% acknowledgements and thanks
% @author Tobias Weber <tweber@ill.fr>
% @date july-2021
% @license see 'LICENSE' file
%

\chapter*{Acknowledgements and Thanks}
\addcontentsline{toc}{chapter}{Acknowledgements and Thanks}

I wish to thank both my thesis advisors, Prof. Dr. Christian Icking and Dr. Lihong Ma, for accepting
and supporting this project, for their advice and our regular discussions, the friendly atmosphere, 
as well as for always being available for questions.

Equal thanks go to Dr. Martin B\"ohm, leader of the spectroscopy group at the ILL and instrument
scientist at the \textit{Thales} spectrometer \cite{thales}, to Yannick Le Goc (computer scientist at the ILL's instrument
control and electronics group), as well as to Dr. Paul Steffens (\textit{Thales} instrument scientist) for our
regular discussions about various triple-axis automatisation projects, our experiments and the
instrumental aspects. I furthermore wish to thank Jérôme Locatelli from the ILL's instrument control group for
all the help with \textit{NOMAD} \cite{Mutti2011, web_NOMAD}, the instrument control system at the ILL, and the technical discussions
about its internal details. 
Thanks to Dr. Paolo Mutti, the head of both the scientific computing and the instrument control group 
at the ILL, for supporting this project and for the great atmosphere and working environment there.
I thank Alain Filhol for his help with creating \textit{MacOS} application packages and build scripts.

Also not to forget the continuing support from my former advisor for past theses, Prof. Dr. Peter B\"oni,
as well as from Prof. Dr. Christian Pfleiderer, Prof. Dr. Markus Garst, and Dr. Marc Janoschek in all our ongoing projects.

Financial support by the Institut Laue-Langevin via its \textit{P\^ole Formation} is gratefully acknowledged,
and I wish to thank Carole Fauchier for all the administrative help.

Last but not least also many thanks to my family and to Soilhat for always being there for me!
